### 2024-08-26 parts/correction.tex
counterexamples

@@

\AP Remark that our second goal is at odds with our restriction to polynomials
with coefficients in $\Rel$, because of the polynomial $X(X-1)/2$
(\cref{q-polynomial-n-poly:ex}). However, we argue that up to a suitable change
of basis for the ring of polynomials, all coefficients will actually be in
$\Nat$. This relatively simple trick works by considering as ``basic
polynomials" the \intro{binomial function} of the form $\binom{X - \ell}{k}$,
where $\ell$ and $k$ are natural numbers. We extend this to \intro{natural
binomial functions} that are obtained by considering $\Nat$-linear combinations
of products of \kl{binomial functions}, that is, we consider functions that
have the following shape: $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i \prod_{j =
1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is
defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial. For
this reason, we also introduce its ``polynomial counterpart" $\intro*\pbinom{X
- \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
- p_{i,j}}{k_{i,j}}$.

@@

Let us also address the fact that the distinction between \kl{natural binomial
functions} and \kl{natural binomial polynomials} vanishes when we are
interested in polynomials. As a consequence, to a \kl{natural binomial
function} corresponds exactly one \kl{natural binomial polynomial} that
coincides with it on sufficiently large inputs (\cref{natural-binomial-polynomial:lem})
which allows us to conclude that \kl{$\Nat$-rational polynomials} are in
$\CorrectPoly$ (\cref{n-rat-correct:lem}).

@@

@@

@@

@@

\begin{corollary}
    \label{decide-rat-poly-npoly:cor}
    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
    \begin{enumerate}
        \item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
        \item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
        \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
        \item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
            and for all $\vec{x} \in \Nat^{\vec{X}}$,
            $P(\vec{x})$ is a multiple of $\alpha$.
    \end{enumerate}
    In particular, the properties are decidable.
\end{corollary}
\begin{proof}
    First, let us remark that 
    \cref{rat-npoly-0:item} $\iff$ \cref{rat-npoly-1:item}
    because \kl{natural binomial polynomials} are increasing.
    Furthermore, \cref{rat-npoly-3:item} $\implies$ \cref{rat-npoly-2:item}
    because $\NSF \subseteq \NPoly$.

    Then, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
    if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
    polynomial}. But in particular, there exists a polyregular function $f$
    that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
    post-compose this function by the one that maps $a^{k \alpha + r}$ to
    $a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
    \kl{represented} by an \kl{$\Nat$-polyregular function}.

    We also know that \cref{rat-npoly-2:item} $\implies$
    \cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
    polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
    \CorrectPoly$ by \cref{corrected-version:thm}.

    As a consequence, we have proven that \cref{rat-npoly-0:item} $\iff$
    \cref{rat-npoly-1:item} and \cref{rat-npoly-2:item} $\iff$
    \cref{rat-npoly-3:item} $\iff$ \cref{rat-npoly-4:item}.

    Finally, \cref{rat-npoly-1:item} $\implies$ \cref{rat-npoly-4:item} is
    exactly \cref{natural-binomia-correct:lem}. And \cref{rat-npoly-3:item}
    $\implies$ \cref{rat-npoly-1:item} is obtained
    by combining \cref{n-sf-combinatorics:lem} and \cref{natural-binomial-polynomial:lem}.
\end{proof}

@@

\begin{corollary}
    \label{integer-binomial-polynomials:cor}
    Let $P \in \Rat[\vec{X}]$. Then, $P$ is a \kl{$\Rel$-rational polynomial}
    if and only if $P$ is an \kl{integer binomial polynomial}.
\end{corollary}
\begin{proof}
    Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
    \kl{star-free $\Nat$-polyregular functions}.
    One can apply \cref{n-sf-combinatorics:lem}
    to conclude that $P$ coincides with an 
    \kl{integer binomial polynomial} over large enough inputs,
    hence that $P$ is such a polynomial.
\end{proof}

>>>>

\subsubsection{From $\Nat$-polyregular functions to polynomials}

@@

\AP Remark that our second goal is at odds with our restriction to polynomials
with coefficients in $\Rel$, because of the polynomial $X(X-1)/2$
(\cref{q-polynomial-n-poly:ex}). However, we argue that up to a suitable change
of basis for the ring of polynomials, all coefficients will actually be in
$\Nat$. This relatively simple trick works by considering as ``basic
polynomials" the \intro{binomial function} of the form $\binom{X - \ell}{k}$,
where $\ell$ and $k$ are natural numbers. We extend this to \intro{natural
binomial functions} that are obtained by considering $\Nat$-linear combinations
of products of \kl{binomial functions}, that is, we consider functions that
have the following shape: $f(x_1, \dots, x_k) = \sum_{i = 1}^n n_i \prod_{j =
1}^k \binom{x_j - p_{i,j}}{k_{i,j}}$. Beware that $\binom{X - \ell}{k}$ is
defined to be $0$ when $X \leq \ell$, and is therefore not a polynomial.

@@

@@

\subsubsection{From polynomials to $\Nat$-polyregular functions}

@@

\subsubsection{From $\Rel$ to $\Rat$ and $\Real$}

@@

For
this reason, we also introduce its ``polynomial counterpart" $\intro*\pbinom{X
- \ell}{k}$ defined as $(X - \ell) \cdots (X - \ell - k) / k!$, that we call a
\intro{binomial monomial}, and introduce \intro{natural binomial polynomials}
as $\Nat$-linear combinations of products of \kl{binomial monomials}, i.e., of
the shape: $P(X_1, \dots, X_k) = \sum_{i = 1}^n n_i \prod_{j = 1}^k \pbinom{X_j
- p_{i,j}}{k_{i,j}}$. 
Let us also address the fact that the distinction between \kl{natural binomial
functions} and \kl{natural binomial polynomials} vanishes when we are
interested in polynomials. As a consequence, to a \kl{natural binomial
function} corresponds exactly one \kl{natural binomial polynomial} that
coincides with it on sufficiently large inputs (\cref{natural-binomial-polynomial:lem})
which allows us to conclude that \kl{$\Nat$-rational polynomials} are in
$\CorrectPoly$ (\cref{n-rat-correct:lem}).

\begin{conjecture}
    Let $P \in \Real[\vec{X}]$. Then $P \in \CorrectPoly$ if and only if $P$ 
    there exists $K$ such that
    $\restr{P}{\nu}$ is a \kl{natural binomial polynomial} \kl{in base $K$}
    for all partial valuations $\nu \colon \vec{X} \topartial \set{0, \dots, K}$.
    That is, for all $\nu$, there exists $\alpha_i \geq 0$ and $d_{i,j} \geq 0$ such that
    $\restr{P}{\nu} = \sum_i \alpha_i \prod_j \pbinom{X_j - K}{d_{i,j}}$.
\end{conjecture}

@@

\begin{corollary}
    \label{decide-rat-poly-npoly:cor}
    Let $P \in \Rat[\vec{X}]$ be a polynomial with \emph{rational}
    coefficients and let $\alpha$ be the smallest number in $\Nat_{\geq 1}$
    such that $\alpha P \in \Rel[\vec{X}]$. Then, the following are equivalent:
    \begin{enumerate}
        %\item \label{rat-npoly-0:item} $P$ is a \kl{natural binomial polynomial} such that $P(0,\dots,0) \geq 0$,
        %\item \label{rat-npoly-1:item} $P$ is a \kl{non-negative} \kl{natural binomial polynomial},
        \item \label{rat-npoly-2:item} $P$ is a \kl{$\Nat$-rational polynomial},
        \item \label{rat-npoly-3:item} $P$ is \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item \label{rat-npoly-4:item} $\alpha P \in \CorrectPoly$,
            and for all $\vec{x} \in \Nat^{\vec{X}}$,
            $P(\vec{x})$ is a multiple of $\alpha$.
    \end{enumerate}
    In particular, the properties are decidable.
\end{corollary}
\begin{proof}
    First, \cref{rat-npoly-4:item} $\implies$ \cref{rat-npoly-3:item}, because
    if $\alpha P \in \CorrectPoly$, then it is an \kl{$\Nat$-rational
    polynomial}. But in particular, there exists a polyregular function $f$
    that represents $\alpha P$ with outputs in $\set{ a }$. Now, let us
    post-compose this function by the one that maps $a^{k \alpha + r}$ to
    $a^{k}$, which is a \kl{polyregular function}. We conclude that $P$ is
    \kl{represented} by an \kl{$\Nat$-polyregular function}.

    We also know that \cref{rat-npoly-2:item} $\implies$
    \cref{rat-npoly-4:item}, since if $P$ is an \kl{$\Nat$-rational
    polynomial}, then so is $\alpha P$, and in particular $\alpha P \in
    \CorrectPoly$ by \cref{corrected-version:thm}.
\end{proof}

@@

It is a longstanding result that polynomials with $\Rel$-output are exactly
\kl{integer binomial polynomials} \cite{POLYA1915,CACHA1996}. However, its
generalization to $\Nat$ is not immediate. First, having non-negative
coefficients depends on a choice of basis, and second, the non-negativity is
not enough to ensure that the polynomial is a \kl{$\Nat$-rational polynomial}.

\begin{example}
    The polynomial $\pbinom{X - 2}{2} + 1$ is a \kl{non-negative} \kl{natural binomial polynomial}
    but in the standard basis, it is written 
    $\pbinom{X}{2} - \pbinom{X}{1} + 2$, which has a \kl{negative} coefficient.
\end{example}

\begin{example}
    The polynomial
    $Q(X,Y,Z) \defined \pbinom{X - 4}{1} \pbinom{Y}{1} \pbinom{Z}{1}
    + 8 \pbinom{Y}{2} + 8 \pbinom{Z}{2} + 4$
    is a \kl{non-negative} \kl{natural binomial polynomial},
    but cannot be computed by a \kl{star-free $\Nat$-polyregular function}.
\end{example}

\begin{corollary}
    \label{integer-binomial-polynomials:cor}
    Let $P \in \Real[\vec{X}]$. Then, the following propreties are equivalent:
    \begin{enumerate}
        \item \label{int-bin-0:item} $P$ is an \kl{integer binomial polynomial},
        \item \label{int-bin-1:item} $P$ \kl{represented} by a \kl{star-free $\Nat$-polyregular function},
        \item \label{int-bin-2:item} $P(\Rel^k) \subseteq \Rel$.
    \end{enumerate}
\end{corollary}
\begin{proof}
    Let $f = f_+ - f_-$ where both $f_+$ and $f_-$ are 
    \kl{star-free $\Nat$-polyregular functions}.
    One can apply \cref{n-sf-combinatorics:lem}
    to conclude that $P$ coincides with an 
    \kl{integer binomial polynomial} over large enough inputs,
    hence that $P$ is such a polynomial.
\end{proof}

### 2024-08-26 parts/prelim.tex
test
counterexamples

\AP Let us briefly recall that one of the equivalent definitions of a
\intro{$K$-rational series} (where $K$ is a ring), and refer to \cite{BERE10}
for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
to $K$ are the functions computed by \intro{$K$-weighted automata} which are
non-deterministic finite automata with transitions labelled by elements of $K$,
and whose output on a given word is the sum over all accepting paths of the
products of the weights along the path. For instance, the function $w \mapsto
2^{\card{w}}$ is computed by an \kl{$\Nat$-weighted automaton} with a single
state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
series}.

@@

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $k \in \Nat$. The set $\intro*\ZPoly[k]$ of polyregular
    \intro{$\Rel$-polyregular functions} of degree at most $k$,
    is the set of functions $f \colon \Sigma^* \to \Rel$
    such that
    there exists a finite monoid $M$,
    a morphism $\mu \colon \Sigma^* \to M$,
    and a function $\pi \colon M^{k+1} \to \Rel$
    satisfying for all $w \in \Sigma^*$:
    \begin{equation*}
        f (w) = \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_{k+1}} \pi(\mu(u_1), \dots, \mu(u_{k+1}))
        \quad .
    \end{equation*}
    We call $\pi$ the \intro{production function} of $f$.
    If the function $\pi$ has codomain $\Nat$,
    then $f$ is \intro{$\Nat$-polyregular} of degree at most $k$,
    i.e., $f \in \intro*\NPoly[k]$.
    If the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \kl{star-free $\Rel$-polyregular}
    ($\intro*\ZSF[k]$), resp. \kl{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
\end{definition}

\AP We complete \cref{nat-rel-poly:def} by letting $\intro*\NPoly \defined
\bigcup_{k \in \Nat} \intro*\NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
$\ZSF$. In order to illustrate these definitions, let us provide an example of
a \kl{$\Nat$-polyregular function} computed using a finite monoid in
\cref{size-of-word-nsf:ex}. Let us also introduce in
\cref{q-polynomial-n-poly:ex} a function that will appear frequently in this
paper as an example of \emph{division} computed by a \kl{$\Nat$-polyregular
function}.

@@

>>>>

\AP Let us briefly recall that one of the equivalent definitions of a
\intro{$K$-rational series} (where $K$ is a ring), and refer to \cite{BERE10}
for a comprehensive introduction. The \kl{$K$-rational series} from $\Sigma^*$
to $K$ are the functions computed by \kl{$K$-weighted automata} which are
non-deterministic finite automata with transitions labelled by elements of $K$,
and whose output on a given word is the sum over all accepting paths of the
products of the weights along the path. For instance, the function $w \mapsto
2^{\card{w}}$ is computed by an \kl{$\Nat$-weighted automaton} with a single
state, and a loop labelled by $2$, and is therefore an \kl{$\Nat$-rational
series}. We use $\intro*\NRat$ and $\intro*\ZRat$ to denote respectively
the set of \kl{$\Nat$-rational series} and the set of \kl{$\Rel$-rational
series}.

@@

\begin{definition}[$\Rel$-polyregular functions {\cite{CDTL23}}]
    \label{nat-rel-poly:def}
    Let $k \in \Nat$. The set $\intro*\ZPoly[k]$ of polyregular
    \intro{$\Rel$-polyregular functions} of degree at most $k$,
    is the set of functions $f \colon \Sigma^* \to \Rel$
    such that
    there exists a finite monoid $M$,
    a morphism $\mu \colon \Sigma^* \to M$,
    and a function $\pi \colon M^{k+1} \to \Rel$
    satisfying for all $w \in \Sigma^*$:
    \begin{equation*}
        f (w) = \pi^\dagger (w) \defined
        \sum_{w = u_1 \cdots u_{k+1}} \pi(\mu(u_1), \dots, \mu(u_{k+1}))
        \quad .
    \end{equation*}
    We call $\pi$ the \intro{production function} of $f$.
    If the function $\pi$ has codomain $\Nat$,
    then $f$ is \intro{$\Nat$-polyregular} of degree at most $k$,
    i.e., $f \in \intro*\NPoly[k]$.
    If the monoid $M$ is \kl(monoid){aperiodic}
    then
    the function $f$ is \intro{star-free $\Rel$-polyregular}
    ($\intro*\ZSF[k]$), resp. \intro{star-free $\Nat$-polyregular} ($\intro*\NSF[k]$).
\end{definition}

\AP We complete \cref{nat-rel-poly:def} by letting $\reintro*\NPoly \defined
\bigcup_{k \in \Nat} \NPoly[k]$, and similarly for $\ZPoly$, $\NSF$, and
$\ZSF$. In order to illustrate these definitions, let us provide an example of
a \kl{$\Nat$-polyregular function} computed using a finite monoid in
\cref{size-of-word-nsf:ex}. Let us also introduce in
\cref{q-polynomial-n-poly:ex} a function that will appear frequently in this
paper as an example of \emph{division} computed by a \kl{$\Nat$-polyregular
function}.

@@

\AP In full generality, \cref{size-of-word-nsf:ex,q-polynomial-n-poly:ex} can be generalized to demonstrate that functions
of the following functions can be \kl{represented} by a \kl{commutative}
\kl{star-free $\Nat$-polyregular function}: $x \mapsto \binom{x - p}{k}$ for $p
\in \Nat$ and $k \in \Nat$, where $\binom{n}{k}$ is the binomial coefficient,
that is the number of ways to choose $k$ elements among $n$ elements.
